---
title: "Analysis of ToothGrowth Data"
output: pdf_document
---
This work is done for the course project for Coursera Statistical Inference. 
The project uses the ToothGrowth dataset in R. 

# Introduction
Load the data into R and examine the file. Load ggplot library for plots. 

```{r}
library(ggplot2)
library(plyr)
data(ToothGrowth)
head(ToothGrowth)
str(ToothGrowth)
```
Three variables that are len (numeric), supp (factor), and dose (numeric).
But dose should be a factor, so convert it from a numeric to factor. 
```{r}
ToothGrowth$dose <-as.factor(ToothGrowth$dose)
```

# Exploratory Data Analysis
Create a boxplot which will display the mean tooth length, the interquartile range, 
and total range for the data, by both dose group (0.5, 1, or 2), and separated by
supplement group (OJ or VC). This will best show the potentially difference between
the groups.

```{r, echo=FALSE, fig.height=3, fig.width=5}
a <- ggplot(ToothGrowth, aes(x=dose, y=len))
a + geom_boxplot(aes(fill = supp)) +
    theme_bw() +
    ggtitle("Tooth length by Treatment and Dose")
```
Dose of both types of supplement appear to have an effect on tooth length. 

# Data Summary
Summarize the tooth length data, finding the overall mean, as well as the mean by 
group (group being dose and supplement group).
```{r}
### mean of columns
mean(ToothGrowth$len)
### sum and mean of len by supp and dose
ddply(ToothGrowth, .(supp, dose), summarise, mean=mean(len), sum=sum(len))
```

# Confidence intervals and hypothesis test
## Test Tooth Length by Supplement
Run a t-test to assess if supplement leads to significant changes in tooth growth. 
```{r}
t.test(len ~ supp, paired = F, var.equal = F, data = ToothGrowth)
```
P-value is greater than 0.05, therefore we can not reject the null hypothesis that
this is not difference between the two supplement groups on tooth length. We can 
further see evidence of this given the 95% confidence intervals contain 0. 

## Test Tooth Length by Dose
In order to examine the effects of dose on tooth length, we first need to create 
dummay variables for dose in order to perform a t-test (this is because dose has
three levels instead of two).
```{r}
dose12 <- subset(ToothGrowth, dose %in% c(0.5, 1.0))
dose13 <- subset(ToothGrowth, dose %in% c(0.5, 2.0))
dose23 <- subset(ToothGrowth, dose %in% c(1.0, 2.0))
```
Run t-tests comparing dose 0.5 to 1, 0.5 to 2 and 1 to 2. 
```{r}
t.test(len ~ dose, paired = F, var.equal = F, data = dose12)
t.test(len ~ dose, paired = F, var.equal = F, data = dose13)
t.test(len ~ dose, paired = F, var.equal = F, data = dose23)
```
Each of those t-test results i p-value below 0.05, which means we can reject the
null hypothesis that there is no difference between the groups, and instead accept
the alternative hypothesis. Furthermore, we can also see in the output that the 
95 % confidence intervals for each t-test do not contain 0. Therefore it seems that 
dose is having an effect on tooth length. 

## Test Tooth Length by Dose and Supplement
Subset the data by dose. 
```{r}
dose5 <- subset(ToothGrowth, dose == 0.5)
dose10 <- subset(ToothGrowth, dose == 1.0)
dose20 <- subset(ToothGrowth, dose == 2.0)
```
Run t-tests within each subset of dose for the relationship between tooth
length and supplement. 
```{r}
t.test(len ~ supp, paired = F, var.equal = F, data = dose5)
t.test(len ~ supp, paired = F, var.equal = F, data = dose10)
t.test(len ~ supp, paired = F, var.equal = F, data = dose20)
```
P-values were below 0.05 in the dose subests of 0.5 and 1.0, but not for 2.0. 
This means we can reject the null hypothesis for the for supplement type and dose 
effect tooth length together at doses 0.5 and 1.0 but not at 2.0. 

# Conclusions
We see that there are difference in tooth length by supplement at specific doses,
but not all doses. We see for doses 0.5 and 1.0 OJ results in significantly (p<0.05)
longer tooths than VC. However, this is not true for doses of 2.0 where there was
not a statistically significant difference in tooth length between the groups.

