---
title: "Exponential Distribution and the Cental Limit Theory"
output: pdf_document
---

# Overview

For this part of the course project for Coursera Statistical Inference, I am 
investigating the exponential distribution in R and comparing it with the 
Central Limit Theorem. I will be answering three questions comparing sample 
means and variance to their theortetical counterparts. I will be simulating the
exponential distribution in R with rexp(n, lambda) where lambda is the rate parameter. 

Lambda will set to 0.2 for all of the simulations. I will be taking the 
distribution of averages of 40 exponentials, and performing 1,000 simulations. 

First load reuqired libraries for plotting.
```{r}
library(ggplot2)
```

# Simulations
Set parameters for simulations using guidelines stated above. Because we are 
running simulations, we will also set the seed to ensure the data is reproducible. 
```{r}
lambda <- 0.2
n <- 40 
number_simulations <- 1000
set.seed(12345)
```
Mean of each random exponential is calculated for the thousand simulations and 
stored in a dataframe called "means". This is done using the values defined 
above in the calculations. 
```{r}
### run simulation getting the mean for 1000 simulations
means <- NULL
for (i in 1:number_simulations) means = c(means, mean(rexp(n, lambda)))
means <- as.data.frame(means)
```
```{r, echo=F, fig.height=3, fig.width=4}
a <- ggplot(means, aes(x=means))
a + geom_histogram(alpha = .50, binwidth=.2, colour="black", fill="white", 
                   aes(y = ..density..)) +
    ggtitle("Histogram of the means from\nthe 1000 simulations") +
    theme_bw()
```

# Sample Mean versus Theoretical Mean
Theoretical mean of an exponetial distribution is 1/lambda. This is compared to 
the mean of the simulation results. 
```{r}
### theoretical mean 
theor_mean <- 1/lambda
### simulation mean
sample_mean <- mean(means$means)
print(c(theor_mean, sample_mean))
```
My sample mean (4.972) is very close the the theoretical mean (5).

```{r, echo=FALSE, fig.height=3, fig.width=4}
a <- ggplot(means, aes(x=means))
a + geom_histogram(alpha = .50, binwidth=.2, colour="black", fill="white", 
                   aes(y = ..density..)) +
    geom_vline(aes(xintercept=theor_mean)) +
    geom_vline(aes(xintercept=sample_mean, colour = "red")) +
    theme_bw() +
    ggtitle("Histogram of the means from\nthe 1000 simulations\nBlack line is theoretical mean\nRed line is sample mean")
```

# Sample Variance versus Theoretical Variance
Theoretical sampling variance is 1/(n*lambda^2), where n in this case is 
equal to 40. This is compared to the sample variance for the simulated results. 
```{r}
### theoretical variance of an exponetial distribution is 1/(n*lambda^2)
theor_var <-  1/(n * lambda^2)
### simulation variance
sample_var <- var(means$means)
print(c(theor_var, sample_var))
```
My theoretical variance (0.595) is close to the theoretical variance (0.625).

```{r, echo=FALSE, fig.height=3, fig.width=4}
a <- ggplot(means, aes(x=means))
a + geom_histogram(alpha = .50, binwidth=.2, colour="black", fill="white", 
                   aes(y = ..density..)) +
    geom_vline(aes(xintercept=theor_mean-theor_var)) +
    geom_vline(aes(xintercept=theor_mean+theor_var)) +
    geom_vline(aes(xintercept=sample_mean+sample_var, colour = "red")) +
    geom_vline(aes(xintercept=sample_mean-sample_var, colour = "red")) +
    theme_bw() +
    ggtitle("Histogram of the means from\nthe 1000 simulations\nBlack lines are theoretical variance\nRed lines are sample variance")
```

# Distribution
Create a plot to compare my distribution of simulated means to the normal distribution.

```{r, echo=FALSE, fig.height=3, fig.width=4}
a <- ggplot(means, aes(x=means))
a + geom_histogram(alpha = .50, binwidth=.2, colour="black", fill="white", 
                   aes(y = ..density..)) +
    stat_function(fun = dnorm, arg = list(mean = theor_mean,
                                          sd = sqrt(theor_var))) +
    geom_density(inherit.aes=F, aes(rnorm(n = number_simulations, 
                                          mean = sample_mean,
                                          sd = sqrt(sample_var))), 
                 linetype='dashed', colour = "red") +
    geom_vline(aes(xintercept=theor_mean)) +
    geom_vline(aes(xintercept=sample_mean, colour = "red")) +
    theme_bw() +
    ggtitle("Histogram of the means from\nthe 1000 simulations\nCompared with the normal distribution (black line)")
```

The distribution of my simulation means (dotted red line) is very close to the 
normal distribution (solid black line). Taking a greater number of simulations 
would results in the simulation means distribution falling even closer to that 
of the normal distribution. 

# Conclusions
The distribution of the averages of the exponential distribution is nearly normal.
A exponential random distributoin have a minimum value of zero and a long skew tail.
This distribution looks entirely different to the almost normal distribution 
obtained after taking the average of a thousand simulations. 
Thus I have demonstated that with larger sample size, the distribution of 
averages become closer to normal distribution even using the exponential 
distribution, which is the Central Limit Theorem.
